An iterated Grothendieck construction of semi-cubical types

### 2021-10-25 paper/paper.tex
[Cubical] Start with mkCube
[paper] Whiteboarding with Hugo

\usepackage[allbordercolors={.192157 .76862 .28627}]{hyperref} % Emerald Green
\usepackage{amsthm, amssymb, amsaddr, fancyhdr, bookmark, fontspec, xargs, multirow, polyglossia, luacode, minted, color, soulutf8, tcolorbox, tabularx, tikz-cd, booktabs, setspace, makecell, caption, mathrsfs, xpatch, lmodern}
\usepackage[hmargin=0.4in, vmargin=0.8in, headheight=10pt]{geometry}

@@

% Universes
\newcommand{\sort}[1]{\ensuremath{\mathsf{U}_{#1}}}

@@

@@

% Code listings
\usemintedstyle{tango}

\title{An iterated Grothendieck construction of semi-cubical types}
\author{Ramkumar Ramachandra}
\address{Université de Paris}
\author{Hugo Herbelin}
\address{Institut de Recherche en Informatique Fondamentale}

@@

Our contribution takes a different direction from traditional contributions to the field of HoTT: although it is a proposed model of HoTT, it uses \emph{unicity of identity proofs} (UIP) instead of univalence to finish off the final coherence condition; the two axoims are mutually incompatible.

@@

  \begin{align*}
    obj(\Simplex{}) & := \sq{n}, \text{regarded as a totally-ordered set}   \\
    mor(\Simplex{}) & := \sq{m} \rightarrow \sq{n}, \text{order-preserving}
  \end{align*}
\end{definition}

@@

  \begin{align*}
    obj(\SSet)             & := \text{functors} \; \Simplex{}^\op \rightarrow \Set      \\
    mor(\SSet)             & := \text{natural transforms between the functors}          \\
    \text{composition law} & := \text{the usual composition of natural transformations} \\
  \end{align*}
\end{definition}

@@

\begin{definition}[\Cube{}]
  \begin{align*}
    obj(\Cube{}) & := \I{n}                                           \\
    mor(\Cube{}) & := \delta^\epsilon_i : \I{n + 1} \rightarrow \I{n}
  \end{align*}

@@

\begin{definition}[$\CSet$ in terms of objects and morphisms]
  \begin{align*}
    obj(\CSet) & := X_n                                                   \\
    mor(\CSet) & := X_\lambda, \text{where $\lambda$ is \Cube{}-morphism}
  \end{align*}

@@

@@

\begin{listing}[H]
  \begin{minted}[linenos]{coq}
    Theorem le_trans {p q n} : p <= q -> q <= n -> p <= n.
    (* ... a non-trivial proof ... *)
    Defined.
  \end{minted}
  \caption{↕ definition, when $\leq$ is defined using an inductive approach}
\end{listing}

\begin{listing}[H]
  \begin{minted}[linenos]{coq}
    Definition le_trans {n m p} (Hnm : n <= m) (Hmp : m <= p) : n <= p :=
      fun q (Hqn : le' q n) => Hmp _ (Hnm _ Hqn).
  \end{minted}
  \caption{↕ definition, when $\leq$ is defined using a nested implication}
\end{listing}

@@

\begin{listing}[H]
  \begin{minted}{coq}
  Theorem le_trans_comm4 {n m p} (Hmn : S n <= S m) (Hmp : S m <= S p) :
  ⇓ (Hmn ↕ Hmp) =S (⇓ Hmn) ↕ (⇓ Hmp).
    reflexivity.
  Defined.
  \end{minted}
\end{listing}

\subsection{Raising and lowering operations on \texorpdfstring{\leq}{<=}}
\mint{coq}/Notation "↕" := (Hnm : n <= m) (Hmp : m <= p) : n <= p/
\mint{coq}/Notation "↑" := (Hnm : n <= m) : n <= S m/
\mint{coq}/Notation "↓" := (Hnm : S n <= m) : n <= m/
\mint{coq}/Notation "⇓" := (Hnm : S n <= S m) : n <= m/
\mint{coq}/Notation "⇑" := (Hnm : n <= m) : S n <= S m/
\subsection{Yoneda.v}
\subsection{RewLemmas.v}
\subsection{Notes on SProp}
\subsection{Cubical.v}

@@

@@

  \bibitem[Her]{Herbelin15}
  Herbelin, H. (2015). A dependently-typed construction of semi-simplicial types. \textit{Mathematical Structures in Computer Science, 25}(5), 1116-1131.
\end{thebibliography}

>>>>

\usepackage[allbordercolors={.192157 .76862 .28627}]{hyperref} % Emerald Green
\usepackage{amsthm, amssymb, amsmath, mathtools, amsaddr, fancyhdr, bookmark, fontspec, xargs, multirow, polyglossia, luacode, minted, color, soulutf8, tcolorbox, tabularx, tikz-cd, booktabs, setspace, makecell, caption, mathrsfs, xpatch, lmodern}
\usepackage[hmargin=0.4in, vmargin=0.8in, headheight=10pt]{geometry}

@@

% Universes, obj, mor
\newcommand{\sort}[1]{\ensuremath{\mathsf{U}_{#1}}}

@@

% Objects and morphisms
\newcommand{\obj}{\ensuremath{\mathsf{obj}}}
\newcommand{\mor}{\ensuremath{\mathsf{mor}}}

@@

% Code listings
\usemintedstyle{tango}
\setminted{escapeinside=~~}
\setmintedinline{escapeinside=~~}

% Fonts
\setmainfont{Avenir Next}
\setmonofont{Source Code Pro}
\newfontfamily\mainfont{Avenir Next}
\newfontfamily\caslon{Adobe Caslon Pro}
\defaultfontfeatures{Ligatures=TeX, Scale=MatchUppercase}

\title{An iterated Grothendieck construction of semi-cubical types}
\author{Hugo Herbelin}
\address{Institut de Recherche en Informatique Fondamentale}
\author{Ramkumar Ramachandra}
\address{Université de Paris}

@@

Our contribution uses the indexed approach to constructing semi-cubical types, as opposed to CCHM~\cite{Cohen16} in 2016, which takes a fibered approach. We will expound on this further in the Grothendieck construction section. We also use \emph{unicity of identity of proofs} (UIP) to finish off the last coherence condition, and we have dedicated a section to discussing it in various settings.

\begin{table}[H]
  \begin{tabularx}{\linewidth}{p{.3\linewidth}|p{.3\linewidth}|p{.3\linewidth}}
    \toprule
            & Set theory & Type theory \\
    \midrule
    Fibered & CCHM       &             \\
    \midrule
    Indexed &            & Our work    \\
    \bottomrule
  \end{tabularx}
\end{table}

@@

  \begin{align*}
    \obj(\Simplex{}) & := \sq{n}, \text{regarded as a totally-ordered set}   \\
    \mor(\Simplex{}) & := \sq{m} \rightarrow \sq{n}, \text{order-preserving}
  \end{align*}
\end{definition}

@@

  \begin{align*}
    \obj(\SSet)            & := \text{functors} \; \Simplex{}^\op \rightarrow \Set      \\
    \mor(\SSet)            & := \text{natural transforms between the functors}          \\
    \text{composition law} & := \text{the usual composition of natural transformations} \\
  \end{align*}
\end{definition}

@@

\begin{definition}[\Cube{}]
  \begin{align*}
    \obj(\Cube{}) & := \I{n}                                           \\
    \mor(\Cube{}) & := \delta^\epsilon_i : \I{n + 1} \rightarrow \I{n}
  \end{align*}

@@

\begin{definition}[$\CSet$ in terms of objects and morphisms]
  \begin{align*}
    \obj(\CSet) & := X_n                                                   \\
    \mor(\CSet) & := X_\lambda, \text{where \lambda is \Cube{}-morphism}
  \end{align*}

@@

\section{Unicity of identity proofs}
We dedicate this section to discussing UIP in different settings, and supply intuitions into this notion.

UIP is a flavor of proof-irrelevance:

\begin{align*}
  \forall x y, \forall p q : x = y, p = q
\end{align*}

which is to say that any two proofs of \emph{equality} of the same two types are \emph{equal}. In other words, the proofs cannot be distinguished from one another.

Proof irrelevance is an inherent part of set theory and first-order logic, although the property cannot be stated in the language of natural deductions. Intuititively, what it means is that a tree of deductions converges to a result, with no memory of the proof route; the tree itself is forgotten once the final result is obtained.

One way of formalising set theory in type theory is via Aczel sets [ref]. It can be written down as an inductive as follows:

\begin{listing}[H]
  \begin{minted}[mathescape]{coq}
    Inductive SET : Type :=
      node : \forall A : Type, (A -> SET) -> SET
      ~$=_\textrm{SET}$~ : ...
  \end{minted}
\end{listing}

[draw tree structure of set]

Then, proof irrelevance can be stated as:

\begin{align*}
  \forall x y, \forall p q : x =_{\textrm{SET}} y, p = q
\end{align*}

We use UIP in the final coherence condition in \mintinline{coq}{cohbox} and \mintinline{coq}{cohcube}. If we weren't to use UIP in \mintinline{coq}{cohbox}, then we'd have something similar to:

\begin{listing}[H]
  \begin{minted}{coq}
    cohbox : subbox ( subbox ... ) = subbox ( subbox ... )
    coh2box : cohbox ... = cohbox ...
    ...
    cohNbox : cohPredNbox ... = cohPredNbox ...
  \end{minted}
\end{listing}

In attempting to write the proof without UIP, we'd have a relation between \mintinline{coq}{cohNbox} and \mintinline{coq}{cohPredNbox}, which would be very difficult to prove.

@@

\begin{listing}[H]
  \begin{minted}[linenos]{coq}
    Theorem le_trans {p q n} : p <= q -> q <= n -> p <= n.
    (* ... a non-trivial proof ... *)
    Defined.
  \end{minted}
  \caption{when $\leq$ is defined using an inductive approach}
\end{listing}

\begin{listing}[H]
  \begin{minted}[linenos]{coq}
    Definition le_trans {n m p} (Hnm : n <= m) (Hmp : m <= p) : n <= p :=
      fun q (Hqn : le' q n) => Hmp _ (Hnm _ Hqn).
  \end{minted}
  \caption{when $\leq$ is defined using a nested implication}
\end{listing}

@@

\begin{listing}[H]
  \begin{minted}{coq}
  Theorem le_trans_comm4 {n m p} (Hmn : S n <= S m) (Hmp : S m <= S p) :
  ~⇓~ (Hmn ~↕~ Hmp) =S (~⇓~ Hmn) ~↕~ (~⇓~ Hmp).
    reflexivity.
  Defined.
  \end{minted}
\end{listing}

\subsection{Raising and lowering operations on \texorpdfstring{\leq}{<=}}
\mint{coq}/~↕~ := (Hnm : n <= m) (Hmp : m <= p) : n <= p/
\mint{coq}/~↑~ := (Hnm : n <= m) : n <= S m/
\mint{coq}/~↓~ := (Hnm : S n <= m) : n <= m/
\mint{coq}/~⇓~ := (Hnm : S n <= S m) : n <= m/
\mint{coq}/~⇑~ := (Hnm : n <= m) : S n <= S m/
\subsection{Yoneda.v}
\subsection{RewLemmas.v}
\subsection{Notes on SProp}
\subsection{Cubical.v}

@@

  \bibitem[CohCoq]{Cohen16}
  Cohen, C., Coquand, T., Huber, S., \& Mörtberg, A. (2016). Cubical type theory: a constructive interpretation of the univalence axiom. arXiv preprint arXiv:1611.02108.

@@

  \bibitem[Her]{Herbelin15}
  Herbelin, H. (2015). A dependently-typed construction of semi-simplicial types. \textit{Mathematical Structures in Computer Science, 25}(5), 1116-1131.

  \bibitem[CoqInCoq]{Barras97}
  Barras, B., \& Werner, B. (1997). Coq in coq. \textit{Available on the WWW.}
\end{thebibliography}
